---
title: "Examples"
author: "Brandie QC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Examples


## Libraries
```{r}
library(tidyverse) #for easy data manipulations 
library(corrplot) #plotting correlations and getting the significance of correlations 

sem <- function(x, na.rm=FALSE) {           #for calculating standard error
  sd(x,na.rm=na.rm)/sqrt(length(na.omit(x)))
} 
```

## Example using left_join for merging data 
```{r}
ann_cens <- read_csv("../input/UCD_annual_census_20240718_corrected.csv") #read in the annual census data
head(ann_cens) #take a peak 

genotype_info <- read_csv("../input/Genotypes_2023_2024.csv") #read in genotype info
head(genotype_info) #take a peak
unique(genotype_info$`Plant Type`) #look at what the different options are for this column (Main_Parent = TM2 or WL2, Other_Parent=other parent pops, F1 = first gen of crossing, F2 = second gen of crossing)

#make sure the names match 
names(ann_cens)
names(genotype_info)
#unique.ID, bed, row, col are in both data frames, but need to rename the column to "col" in the genotype info data frame

ann_cens_genotype_merge <- genotype_info %>% 
  rename(col=column) %>%  #rename to match the annual census data frame 
  left_join(ann_cens) #this will merge the two data frames based on the column names that match between the two 
#this gives an error because the overlapping columns are different types 
str(ann_cens)
str(genotype_info)
#we can see that while bed, row, and col all match in terms of column type between the two data frames, unique.ID is numeric in genotype_info but a character in ann_cens 

#try the merge again fixing the above issue:
ann_cens_genotype_merge <- genotype_info %>% 
  rename(col=column) %>%  #rename to match the annual census data frame 
  mutate(unique.ID=as.character(unique.ID)) %>% #convert unique.ID to be a character instead of numeric to match the annual census 
  left_join(ann_cens) %>%  #this will merge the two data frames based on the column names that match between the two 
  filter(!is.na(`Plant Type`)) #remove NA plant tyeps 
head(ann_cens_genotype_merge) #check that the merge worked 
#signs that it didn't work would be a bunch of NAs in the new columns you've added, which we don't have here so it looks good

#make a boxplot with stem diameter 
ann_cens_genotype_merge %>% 
  ggplot(aes(x=pop.id, y=diam.mm)) +
  geom_boxplot() +
  facet_wrap(~`Plant Type`) #facet by plant type 

#or make separate plots for different plant types 
ann_cens_genotype_merge %>% 
  filter(`Plant Type`=="Main_Parent") %>% 
  ggplot(aes(x=pop.id, y=diam.mm)) +
  geom_boxplot()

ann_cens_genotype_merge %>% 
  filter(`Plant Type`=="Other_Parent") %>% 
  ggplot(aes(x=pop.id, y=diam.mm)) +
  geom_boxplot()
```

## Example of converting Date characters to actual dates 
```{r}
dummy_dates <- tibble(Date=c("10/16/25", "8/16/25", "7/4/25")) #make a data frame with some dates
dummy_dates #these dates are "characters" according to R right now 

dummy_dates %>% 
  mutate(Date_2=mdy(Date)) #convert those dates to the date format - we used mdy here because our dates are in the month/day/year format 
```


## Example for error bars
```{r}
wl2_establishment <- read_csv("../input/WL2_Establishment.csv") #data for survival 3 weeks post-transplant in 2023 at WL2
head(wl2_establishment) #look at first few rows 

wl2_est_summary <- wl2_establishment %>% 
  group_by(bed) %>% #summarize by bed 
  summarise(meanSurv=mean(Establishment, na.rm=TRUE), #calculate the mean of establishment
            semSurv=sem(Establishment, na.rm=TRUE))  #calculate the standard error of establishment using the function at the top of this script ("sem")
wl2_est_summary

wl2_est_summary %>% 
  ggplot(aes(x=fct_reorder(bed, meanSurv), y=meanSurv)) + #fct_reorder just reorders the beds on the x axis according to their mean establishment 
  geom_col(width = 0.7,position = position_dodge(0.75)) + #make a bar plot 
  geom_errorbar(aes(ymin=meanSurv-semSurv, ymax=meanSurv+semSurv), #add error bars 
                width=.2, position = position_dodge(0.75)) + #give the error bars some properties to match the bars
  theme_classic() + 
  labs(x="Bed", y="Establishment") #label the x and y axis 


## make the same plot as above but filter first
wl2_est_summary %>% 
  filter(bed=="A" | bed=="B") %>% #only keep beds A and B, you could do position == "Top" for example 
  ggplot(aes(x=fct_reorder(bed, meanSurv), y=meanSurv)) +
  geom_col(width = 0.7,position = position_dodge(0.75)) + 
  geom_errorbar(aes(ymin=meanSurv-semSurv,ymax=meanSurv+semSurv),width=.2, position = 
                  position_dodge(0.75)) + 
  theme_classic() +
  labs(x="Bed", y="Establishment", title="Top") #add a title to indicate what position you filtered by 
```

1.  Test for a correlation between TDR and MetStation when the data overlaps

    1.  Correlation test with mean MetStation soil moisture for summer 2024 and mean TDR for each Bed for 2024

    2.  Calculate the difference from the Metstation sensor to see how that varies over time and across positions of TDR measurements (can make figures where the y axis is the difference) 
    
```{r}
wl2_cors <- wl2_establishment %>% 
  select(GrwSsn_GD_Recent, GrwSsn_GD_Historical) #make a dataframe with just the columns you want to correlate

cor.norm = cor(wl2_cors) #test correlations among the traits
cor.norm
cor.sig <- cor.mtest(wl2_cors, method = "pearson") #getting the p-values for correlations 
cor.sig
```

```{r}
test_metdata <- tibble(Date=c("10/16/25", "8/16/25", "7/4/25"), #tibble allows you to make new data frames 
                    MetStation_moisture=c(0, 0, 0.2))
test_metdata

test_TDRdata <- tibble(Date=c("10/16/25", "8/16/25", "7/4/25", "10/16/25", "8/16/25", "7/4/25"),
                       Bed = c("A", "A","A", "B", "B", "B"),
                       Pos = c("Top","Top","Top","Top","Top","Top"),
                       TDR = c(0, 0, 0.2, 0, 0, 0.2))
test_TDRdata

merge_test <- left_join(test_TDRdata, test_metdata) %>% #merge the metstation and TDR data, by using left_join and having the TDR listed first, it will duplicate the metstation data to line up with the TDR data according to the Date column 
  mutate(Moisture_dif = MetStation_moisture-TDR) #calcualte the difference between each of the TDR measurements and the MetStation soil moisture 
merge_test
```

