---
title: "Examples"
author: "Brandie QC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Examples

-   Figure updates:

    -   Sort “over time” figures by date (convert from character to date, Brandie can demonstrate)

    -   Error bars on mean TDR per bed bar plot (Brandie can share some code for this)

    -   Bar plot for mean TDR per position (+ error bars)

    -   Compare Bed\*Pos over time

        -   Example Top of Bed A vs Top of Bed B (similar to the boxplot you made)

## Libraries
```{r}
library(tidyverse)

sem <- function(x, na.rm=FALSE) {           #for calculating standard error
  sd(x,na.rm=na.rm)/sqrt(length(na.omit(x)))
} 
```

## Example of converting Date characters to actual dates 
```{r}
dummy_dates <- tibble(Date=c("10/16/25", "8/16/25", "7/4/25"))
dummy_dates

dummy_dates %>% 
  mutate(Date_2=mdy(Date))
```


1.  Test for a correlation between TDR and MetStation when the data overlaps

    1.  Correlation test with mean MetStation soil moisture for summer 2024 and mean TDR for each Bed for 2024

    2.  Calculate the difference from the Metstation sensor to see how that varies over time and across positions of TDR measurements (can make figures where the y axis is the difference) 
