---
title: "WL2 Garden Correlations"
author: "Brandie QC"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## WL2 Garden Size Correlations 
Test for relationship between stem diameter, canopy area, and height

```{r}
#libraries
library(dplyr)
library(tidyverse)
library(corrplot)
```

## Read in data 
```{r}
WL2_Annual_Census_20241023_corrected <- read_csv("../input/WL2_Annual_Census_20241023_corrected.csv")
WL2_size_survey_20240924_corrected <- read_csv("../input/WL2_size_survey_20240924_corrected.csv")
```

## Merge Data 
```{r}
## check for non-NA canopy cols 
WL2_Annual_Census_20241023_corrected %>% filter(!is.na(overhd.diam))
WL2_size_survey_20240924_corrected %>% filter(!is.na(overhd.diam))
### different rows with canopy info in the different data sheets 

## separate out canopy info from other size data 
WL2_anncens_stemdiam <- WL2_Annual_Census_20241023_corrected %>% select(bed:unique.ID, diam.mm)
WL2_anncens_canopyinfo <- WL2_Annual_Census_20241023_corrected %>% select(bed:unique.ID, overhd.diam:overhd.perp) %>% 
  filter(!is.na(overhd.diam)) #only keep columns with data 

WL2_size_height <- WL2_size_survey_20240924_corrected %>% select(bed:height.cm)
WL2_size_canopyinfo <- WL2_size_survey_20240924_corrected %>% select(bed:unique.ID, overhd.diam:overhd.perp) %>% 
  filter(!is.na(overhd.diam)) #only keep columns with data 

```

### fixing canopy columns
```{r}
WL2_size_canopyinfo <- WL2_size_canopyinfo %>%
  rename(overhd.diam_sizecens = overhd.diam, overhd.perp_sizecens = overhd.perp)

WL2_2024_canopy <- bind_rows(WL2_anncens_canopyinfo, WL2_size_canopyinfo)
dim(WL2_2024_canopy) #286 rows 
WL2_2024_canopy %>% select(unique.ID) %>% distinct() #only 165 plants so definitely some duplicates

#For the duplicates, we want to keep the AnnCens (more recent measurements)
WL2_2024_canopy_noduplicates <- WL2_anncens_canopyinfo %>% 
  full_join(WL2_size_canopyinfo) %>% 
  mutate(overhd.diam=if_else(!is.na(overhd.diam), overhd.diam, #if there's ann cens data, keep it 
                             overhd.diam_sizecens), #if no ann cens data then use size data
         overhd.perp=if_else(!is.na(overhd.perp), overhd.perp, #if there's ann cens data, keep it 
                             overhd.perp_sizecens)) %>% 
  select(bed:overhd.perp)
```

### merge 2024 data
```{r}
Merged_2024 <- WL2_anncens_stemdiam %>% 
  left_join(WL2_size_height) %>% 
  left_join(WL2_2024_canopy_noduplicates)
Merged_2024 %>% filter(!is.na(overhd.diam)) #164 rows with canopy info
# Note that we lost 1 row of canopy info, likely because we didn't have stem diameter for that plant 

Merged_2024$canopy.area <- pi * (Merged_2024$overhd.diam/2) * (Merged_2024$overhd.perp/2)
size_variables <- Merged_2024 %>% select(diam.mm:height.cm, canopy.area)

# save 2024 df
write.csv(Merged_2024, "../input/Merged_2024_sizevariables")
```

## plots for correlations
Brandie changed to point plots since this is the raw data (not means)
```{r}
# height vs canopy area
Merged_2024 %>% 
  ggplot(aes(x=height.cm, y=canopy.area)) +
  geom_point() +
  theme_bw() +
  labs(x="Height", y="Canopy Area", title="Height vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Height_v_Canopy_Area2024.png", width = 20, height = 14, units = "in")

# overhead diameter vs canopy area
Merged_2024 %>% 
  ggplot(aes(x=overhd.diam, y=canopy.area)) +
  geom_point() +
  theme_bw() +
  labs(x="Diameter", y="Canopy Area", title="Overhead Diameter vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Overhead Diameter_v_Canopy_Area2024.png", width = 20, height = 14, units = "in")

# stem diameter vs canopy area
Merged_2024 %>% 
  ggplot(aes(x=diam.mm, y=canopy.area)) +
  geom_point() +
  theme_bw() +
  labs(x="Stem Diameter (mm)", y="Canopy Area", title="Stem Diameter vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Stem Diameter_v_Canopy_Area2024.png", width = 20, height = 14, units = "in")

# overhead perp vs canopy area
Merged_2024 %>% 
  ggplot(aes(x=overhd.perp, y=canopy.area)) +
  geom_point() +
  theme_bw() +
  labs(x="Perpendicular", y="Canopy Area", title="Overhead Perpendicular vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Perpendicular_v_Canopy_Area2024.png", width = 20, height = 14, units = "in")

# stem diameter vs height
Merged_2024 %>% 
  ggplot(aes(x=diam.mm, y=height.cm)) +
  geom_point() + 
  theme_bw() +
  labs(x="Stem Diameter", y="Height", title="Stem Diameter vs Height") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Stem_v_Height2024.png", width = 20, height = 14, units = "in")
```

## 2024 correlation matrix
```{r}

cor_matrix <- cor(size_variables, use = "complete.obs")

png("../output/WL2_2024_Correlation_Matrix.png", width = 800, height = 800)
corrplot(cor_matrix,
         method = "square",
         type = "lower",
         addCoef.col = "black",
         tl.col = "black")
```

## Read in 2025 data
```{r}
WL2_ann_cens_20251028_corrected <- read_csv("../input/WL2_ann_cens_20251028_corrected.csv", 
                                            na=c("", "NA", "N/A"))
WL2_size_survey_20250924_blocksA_F_corrected <- read_csv("../input/WL2_size_survey_20250924_blocksA-F_corrected.csv")
WL2_size_survey_20250929_blocksG_N_corrected <- read_csv("../input/WL2_size_survey_20250929_blocksG-N_corrected.csv")
```

## Merge Data

### Merge size data from 9/24 and 9/29
```{r}
WL2_size_survey_20250924_blocksA_F_corrected %>% filter(!is.na(height.cm))
WL2_size_survey_20250929_blocksG_N_corrected %>% filter(!is.na(height.cm)) #some overlaps between the two 

WL2_size_survey_20250929_blocksG_N_prep <- WL2_size_survey_20250929_blocksG_N_corrected %>% 
  rename_with(~ paste0(.x, "_0929"), height.cm:overhd.perp) %>% 
  select(bed:overhd.perp_0929)

size_survey_2025 <- WL2_size_survey_20250924_blocksA_F_corrected %>% 
  select(bed:overhd.perp) %>% 
  full_join(WL2_size_survey_20250929_blocksG_N_prep) %>% 
  mutate(height.cm=if_else(!is.na(height.cm), height.cm, #if there's 9/24 data, keep it 
                             height.cm_0929),
         overhd.diam=if_else(!is.na(overhd.diam), overhd.diam, #if there's 9/24 data, keep it 
                             overhd.diam_0929), #if no 9/24 then use 9/29 data 
         overhd.perp=if_else(!is.na(overhd.perp), overhd.perp,  
                             overhd.perp_0929)) %>% 
  select(bed:overhd.perp)
```

### Merge size surveys with anncens for height and canopy area 
```{r}
WL2_anncens_heightcanopy_2025 <- WL2_ann_cens_20251028_corrected %>% 
  select(bed:Unique.ID, height.cm=`height. cm`, overhd.diam=`overhd. diam`, overhd.perp=`overhd. perp`) %>% 
  filter(!is.na(overhd.diam))
size_survey_2025_prep <- size_survey_2025 %>% 
  rename_with(~ paste0(.x, "_sizecens"), height.cm:overhd.perp)

#For the duplicates, we want to keep the AnnCens (more recent measurements)
WL2_2025_heightcanopy_noduplicates <- WL2_anncens_heightcanopy_2025 %>% 
  full_join(size_survey_2025_prep) %>% 
  mutate(height.cm=if_else(!is.na(height.cm), height.cm, #if there's ann cens data, keep it
                             height.cm_sizecens),
         overhd.diam=if_else(!is.na(overhd.diam), overhd.diam, #if there's ann cens data, keep it
                             overhd.diam_sizecens), #if no ann cens data then use size data
         overhd.perp=if_else(!is.na(overhd.perp), overhd.perp, #if there's ann cens data, keep it
                             overhd.perp_sizecens)) %>% 
  select(bed:overhd.perp)
```

### Merge all 2025 size
```{r}
WL2_anncens_stemdiam_2025 <- WL2_ann_cens_20251028_corrected %>% select(bed:Unique.ID, diam.mm)
Merged_2025 <- WL2_anncens_stemdiam_2025 %>% 
  left_join(WL2_2025_heightcanopy_noduplicates)

Merged_2025$canopy.area <- pi * (Merged_2025$overhd.diam/2) * (Merged_2025$overhd.perp/2)
size_variables_2025 <- Merged_2025 %>% select(diam.mm:height.cm, canopy.area)

# save 2025 df
write.csv(Merged_2025, "../input/Merged_2025_sizevariables.csv")
```

## plots for 2025 correlations
```{r}
# height vs canopy area
Merged_2025 %>% 
  ggplot(aes(x=height.cm, y=canopy.area)) +
  geom_point() + 
  theme_bw() +
  labs(x="Height", y="Canopy Area", title="Height vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Height_v_Canopy_2025.png", width = 20, height = 14, units = "in")

# overhead perp vs canopy area
Merged_2025 %>% 
  ggplot(aes(x=overhd.perp, y=canopy.area)) +
  geom_point() + 
  theme_bw() +
  labs(x="Overhead Perpendicular", y="Canopy Area", title="Overhead Perpendicular vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Perpendicular_v_CanopyArea_2025.png", width = 20, height = 14, units = "in")

# Overhead diameter vs canopy area
Merged_2025 %>% 
  ggplot(aes(x=overhd.diam, y=canopy.area)) +
  geom_point() + 
  theme_bw() +
  labs(x="Diameter", y="Canopy Area", title="Diameter vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Diameter_v_CanopyArea_2025.png", width = 20, height = 14, units = "in")

# stem diameter vs canopy area
Merged_2025 %>% 
  ggplot(aes(x=diam.mm, y=canopy.area)) +
  geom_point() + 
  theme_bw() +
  labs(x="Stem Diameter", y="Canopy Area", title="Stem Diameter vs Canopy Area") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Stem_v_CanopyArea_2025.png", width = 20, height = 14, units = "in")
# NOTE: stem diameter graph had limited data to work with, not many rows with diam.mm data

# stem diameter vs height
Merged_2025 %>% 
  ggplot(aes(x=diam.mm, y=height.cm)) +
  geom_point() + 
  theme_bw() +
  labs(x="Stem Diameter", y="Height", title="Stem Diameter vs Height") + 
  theme(axis.text.x = element_text(angle = 45, hjust =1, size = 11), text=element_text(size=20))
ggsave("../output/Stem_v_Height2025.png", width = 20, height = 14, units = "in")
```

## 2025 correlation matrix
```{r}

cor_matrix_2025 <- cor(size_variables_2025, use = "complete.obs")

png("../output/WL2_2025_Correlation_Matrix.png", width = 800, height = 800, )
corrplot(cor_matrix_2025,
         method = "square",
         type = "lower",
         addCoef.col = "black",
         tl.col = "black")
```
